---
title: "Meta-analysis"
author: "Amara Santiesteban-Serrano"
date: " r format(Sys.time(), '%d %B, %Y') "
output: 
  html_document:
    df_print: paged
    pdf_document: default
    word_document: default
  editor_options:
    chunk_output_type: inline
---

*Part I*: selecting the studies for the meta-analysis

```{r Packages}
#devtools::install_github("daniel1noble/orchaRd", ref = "main", force = TRUE)
pacman::p_load("ggplot2","tidyr","dplyr","tidyverse","readxl","pastecs", "metagear", "metafor", "orchaRd")

```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("EBImage")
```

```{r Read dataset}
metadata_WOS <- Metaanalysis_ArticleSelection
#metadata_2025 <- Metaanalysis_dataset_NEW2025
Final_dataset <- Metaanalysis_dataset
```

```{r Dealing with duplicates}
#We make sure that R finds titles that are duplicated despite how they are written:
references_cleaned <- metadata_WOS %>%
  mutate(Title = str_to_lower(TITLE),  # Convert to lowercase
         Title = str_replace_all(Title, "[[:punct:]]", ""))  # Remove punctuation

# See duplicated titles
duplicates <- references_cleaned %>%
  dplyr::group_by(Title) %>%
  dplyr::filter(n() > 1)

#Remove duplicates:
references_cleaned <- references_cleaned %>%  
  dplyr::distinct(Title, .keep_all = TRUE)
```

```{r Abstract screening}
effort_distribute(metadata_WOS_2025,
                  reviewers = "Amara",
                  initialize = TRUE,
                  save_split = TRUE)

abstract_screener("effort_Amara3.csv", 
                  "Amara",
                  highlightKeywords = c("fire", "burn", "ectomycorrhiza", 
                                        "mycorrhiza", "fungi"))

#Go for a second round of screening of the DISCARDED abstracts:
data_screening2 <- references_cleaned %>%
  filter(INCLUSION=="NO")

#Proceed to abstract screening: 
effort_distribute(data_screening2,
                  reviewers = "Amara",
                  initialize = TRUE,
                  save_split = TRUE)

abstract_screener("Effort_Amara1.csv", 
                  "Amara",
                  highlightKeywords = c("fire", "burn", "ectomycorrhiza", 
                                        "mycorrhiza", "fungi"))
```

```{r}
#Selected papers for retrieving information (230):
selected_papers <- references_cleaned %>%
  filter(INCLUSION2=="YES")
#Pdf files available (212):
downloaded_papers <- selected_papers %>%
  filter(Pdf_available=="YES")
#Select papers after checking the inclusion criteria (110):
included_papers <- downloaded_papers %>%
  filter(INCLUSION3=="YES")
```

```{r PRISMA flow-chart}
phases <- c("START_PHASE: 573 of studies identified through database searching",
            "564 of studies after duplicates removed",
            "564 of studies with title and abstract screened",
            "EXCLUDE_PHASE: 334 of studies excluded",
            "212 of full-text articles assessed for eligibility",
            "EXCLUDE_PHASE: 98 of full-text excluded, not fitting eligibility criteria",
            "111 of studies included in qualitative synthesis",
            "EXCLUDE_PHASE: # studies excluded, incomplete data reported",
            "final # of studies included in quantitative synthesis (meta-analysis)")
#png("PRISMA_flow_diagram.png", width = 1600, height = 1200, res = 160)
plot_PRISMA(phases)
dev.off()
?plot_PRISMA
```

*Part II*: data analysis

```{r}
ma_dataset <- read_xlsx("Metaanalysis_dataset.xlsx", sheet = "extracted_data")

unique_ids <- ma_dataset %>% 
  distinct(Study_ID, .keep_all = TRUE)

coordinates <- unique_ids %>%
  dplyr::select(Study_ID, Latitude, Longitude)
#write.csv(coordinates, "coordinates.csv")

ggplot(unique_ids, aes(x=fct_infreq(Research_type), fill=Research_type)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Year", y = "Count")
#ggsave("years_meta.png")

ggplot(downloaded_papers, aes(x=fct_rev(fct_infreq(Country)), fill=INCLUSION3)) +
  geom_bar() +
  coord_flip()+
  theme_minimal() +
  labs(x = "Territory", y = "Count")
#ggsave("countries_meta.png")

ggplot(unique_ids, aes(x = Year_of_fire, fill = Year_of_fire)) +
  geom_bar() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

unique_ids %>%
  count(Study_Year, name = "Record_Count") %>%
  arrange((Record_Count))

ggplot(unique_ids, aes(x="", y=Research_type, fill=Research_type)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  scale_fill_brewer(palette="Set1")+
  theme_void()

ggplot(ma_dataset, aes(x=Study_Year, fill=Sample_type)) +
       geom_bar(stat = "count") +
       #scale_fill_manual(values = c("#feb24c", "#f03b20")) +
       theme_minimal()+
  labs(x = "Study year", y = "Count")
```

#################### IMAGES DATA EXTRACTION ####################################

```{r}
figure_scatterPlot(file="scatter_trial.png")
```

```{r Data}
a <- X452_pcoa_10_30
b <- X452_control_10_30
```

```{r Euclidean distances: ORDINATIONS}
## Dt (Distance within treatment group)
t_group <- a %>% 
  #dplyr::filter(group=="2016_10_20") %>% 
  dplyr::select(x,y)

#plot(t_group$x, t_group$y)
D <- dist(t_group, method="euclidean")
Dt <- mean(as.vector(D))
SD_Dt <- sd(as.vector(D))
Dt
SD_Dt


## Dc (Distance within control group)
c_group <- b %>%
  #dplyr::filter(group=="control_2016_10_20") %>% 
  dplyr::select(x,y)

D <- dist(c_group, method="euclidean")
Dc <- mean(as.vector(D))
SD_Dc <- sd(as.vector(D))
Dc
SD_Dc


## Db (Distance between groups)
pairs <- expand.grid(1:nrow(c_group), 1:nrow(t_group))
Db_vals <- apply(pairs, 1, function(idx) {
  sqrt(sum((c_group[idx[1],] - t_group[idx[2],])^2))
})
Db <- mean(Db_vals)
SD_Db <- sd(Db_vals)
Db
SD_Db

```

########################## DATABASE COMPLETED: DATA ANALYSIS ############################

```{r Databases}
means_df <- Metaanalysis_dataset

ordinations_df <- Metaanalysis_dataset #ordination_data
```

```{r COORDINATES}
df_unique <- Metaanalysis_dataset %>%
  distinct(Study_ID, .keep_all = TRUE)

unique_coordinates <- df_unique %>%
  dplyr::select(Study_ID, Latitude, Longitude)
#write.csv(unique_coordinates, "coordinates_metaanalysis.csv")
```

```{r Descriptive statistics}
means_df %>%
  dplyr::count(Forest_type) %>%
  arrange(desc(n))

means_df$effect_size <- ordered(means_df$effect_size, levels=c("hedges_g", "RR", "logRR"))

ggplot(means_df, aes(x=Biome, fill=Biome)) +
  geom_histogram(stat="count")

df_unique %>%
  dplyr::filter(Research_type=="Field_experiment") %>%
  dplyr::count(Guild_studied)

df_counts_cross <- df_unique %>%
  dplyr::count(Research_type, Guild_studied) %>%
  dplyr::arrange(Research_type, desc(n))

df_counts_cross$Research_type <- ordered(df_counts_cross$Research_type, levels=c("Descriptive", "Field_experiment", "Bioassay"))

ggplot(df_counts_cross, aes(x = Research_type, y = n, fill = Guild_studied)) +
  geom_bar(stat = "identity") +
  labs(x = "Study type", y = "Number of studies", fill = "Fungi group") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set2")
#ggsave("research_type.png")

allfungi_descrip <- df_unique %>%
  dplyr::filter(Guild_studied=="All_fungi")

allfungi_counts <- allfungi_descrip %>%
  dplyr::count(Sample_type, Sequencing_type, Sequencing_device, Amplicon)

ggplot(allfungi_counts, aes(x = Sample_type, y = n, fill = Sequencing_type)) +
  geom_bar(stat = "identity") +
  labs(x = "Study type", y = "Number of studies", fill = "Fungi group") +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set2")

ggplot(df_unique, aes(x = Study_Year)) +
  geom_histogram() 
```

1. "Treatments"
   ├─ Burned vs Control
   ├─ Fire_type (Prescribed fire/Wildfire)/ Fire_severity (UB, LS, HS1, HS2)
   └─ 

2. Covariables (categorical or continuous predictors)
   ├─ Soil depth (0–10 cm, 10–20 cm)
   ├─ Time since fire (months)
   ├─ Sample type/sequencing method...
   └─ Climatic region / biome

3. Response variables (what we evaluate)
   ├─ Abundance/richness
   ├─ Diversity α / β
   ├─ % colonization

We are going to use the *metafor* package.

Our database has to have:
*yi*: size of the effect (SMD, log response ratio, etc.)
*vi*: variance of the size of the effect
*moderators*: e.g., ecosystem type, severity, time since fire...

```{r Effect sizes: Hedge's d}

#Although we have already calculated the effect sizes in Excel, we want to try to obtain the same result using the metafor package:

########## STANDARDIZE MEAN DIFFERENCES ---> CALCULATION OF HEDGE'S D:

Effectsizes_df <- means_df %>%
  dplyr::filter(effect_size=="hedges_g")

#intmean = Burnt_value
#intsd = Burnt_SD
#intn = N_burned
#cmean = UB_value
#csd = UB_SD
#cn = N_control

Effectsizes_df$Burnt_SD <- as.numeric(Effectsizes_df$Burnt_SD)
Effectsizes_df$UB_SD <- as.numeric(Effectsizes_df$UB_SD)

Effectsizes <- escalc(measure="SMD", m1i = Burnt_value, sd1i = Burnt_SD, n1i = N_burned,
                      m2i = UB_value, sd2i = UB_SD, n2i = N_control, data=Effectsizes_df) #We got an error saying there was a negative value in one of the SD so we went back to the Excel dataset and corrected it

#Our study with Hedge's d effect sizes is based on *315* entries
```

```{r Random/Mixed Effects models}
random_result <- rma(yi, vi, data = Effectsizes) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(random_result)

#Test for Heterogeneity: we are interested in this! If our p < 0.5 that's a check mark :)
#I^2 (total heterogeneity / total variability):   78.08%
#tau^2 (estimated amount of total heterogeneity): 0.8045 (SE = 0.0934) --> we should consider reporting this in our paper
```

Here we computed all effect sizes together... But we should not mix saprotroph richness with ecm richness with shannon diersity...

```{r Hedges'g effect sizes categorization}

All_fungi_richness <- Effectsizes %>%
  dplyr::filter(Response_var=="Richness")

All_fungi_result <- rma(yi, vi, data = All_fungi_richness) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(All_fungi_result)

ECM_richness <- Effectsizes %>%
  dplyr::filter(Response_var=="ECM_Richness")

ECM_col <- Effectsizes %>%
  dplyr::filter(Response_var == "Colonization_per")
ECM_result <- rma(yi, vi, data = ECM_col) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(ECM_result)

Saprotrophs_richness <- Effectsizes %>%
  dplyr::filter(Response_var=="Saprotroph_richness")

Sporocarps_abun <- Effectsizes %>%
  dplyr::filter(Response_var == "Sporocarp_Abundance")
```

```{r Publication Bias}
funnel(All_fungi_result) #we look for symmetry

##Trim and fill analysis
trimandfill <- trimfill(All_fungi_result)
#Estimated number of missing studies on the right side: 0 (SE = 10.2953)
trimandfill
funnel(trimandfill)

regtest(All_fungi_result)
#Test for Funnel Plot Asymmetry: z = -6.3480, p < .0001 #This SIGNIFICANCE means our funnel plot is ASYMMETRICAL
```

```{r Influence check: OUTLIERS}
infl <- influence(All_fungi_result) #With this function, we check whether there are any studies leveraging the model.
infl$inf #We are going to go back to all those studies with an * in the inf column and check if there was some error
#write.csv(as.data.frame(infl$inf), "influence.csv")
```

```{r Forest plot}
png("ECM_richness_forest_plot.png", width = 2000, height = 3000, res = 300)
forest.rma(All_fungi_result, slab = All_fungi_richness$Study_ID, header="study")
dev.off()
```

But maybe we are putting together things that are way to different... Maybe we should slice our dataset. *Tests of moderators* will tell us if this makes sense!!

```{r Test of Moderators}
#Test of Moderators = Q-BETWEEN. Are there significant differences between levels/groups?
mod.researchtypeq <-rma(yi, vi, mods = ~ factor(Research_type), data= Saprotrophs_alpha)
mod.researchtypeq ###p-val = 0.2268 --> there are NOT significant differences between levels

mod.firetypeq <-rma(yi, vi, mods = ~ factor(Fire_type), data= Saprotrophs_alpha)
mod.firetypeq ### p-val = 0.2014 --> there are NOT significant differences between levels 

mod.firesevq <-rma(yi, vi, mods = ~ factor(Fire_severity), data= Saprotrophs_alpha)
mod.firesevq ### p-val = 0.0006 --> SUPER SIGNIFICANT!!

mod.timesincefireq <- rma(yi, vi, mods = ~ Time_since_fire_months, data=All_fungi_alpha) #continuous moderator
mod.timesincefireq ### p-val < .0001 --> SUPER SIGNIFICANT!!

mod.biomeq <- rma(yi, vi, mods = ~ Biome, data=Saprotrophs_alpha) 
mod.biomeq

mod.foresttypeq <- rma(yi, vi, mods = ~ Forest_type, data=Saprotrophs_alpha) 
mod.foresttypeq ### p-val = 0.7942 --> not significant at all

mod.sampletypeq <- rma(yi, vi, mods = ~ Sample_type, data=Saprotrophs_alpha) 
mod.sampletypeq ### p-val = 0.0248 --> could be something to explore

mod.depthq <- rma(yi, vi, mods = ~ Soil_samples_depth_cm, data=Saprotrophs_alpha) #continuous moderator
mod.depthq ### p-val = 0.1012 ---> Not significant but maybe we should try and make it a CONTINUOUS VARIABLE

mod.responseq <- rma(yi, vi, mods = ~ Response_var, data=Effectsizes) #continuous moderator
mod.responseq ### p-val = 0.0303 --> maybe it matters??

#########################################################################################
# Once we have done the previous models and seen which one of them is significant, now we can run the same model structure WITHOUT the intercept. This will allow us to extract the category means precisely for publication

mod.firesev <-rma(yi, vi, mods = ~ factor(Time_since_fire_months) -1, data= All_fungi_richness) ## -1 = WITHOUT INTERCEPT
mod.firesev
```

```{r Moderators output}
## For publication
mod.grade_table <- coef(summary(mod.firetype))
Firetype_mod <- Effectsizes %>%  ##number of "participants" in each group
  dplyr::group_by(Fire_type) %>%
  dplyr::summarise(nexp = sum(N_burned, na.rm = TRUE),
                   nctrl = sum(N_control, na.rm = TRUE))
Firetype_numcomp <- Effectsizes %>%
  dplyr::count(Fire_type)
Fire.type_finaldf <- cbind(Firetype_mod, Firetype_numcomp[c(2)], mod.grade_table)
```

We can go a step further and carry out a *MULTI-LEVEL MODEL* --> this is the kind of model we use when we have multiple effect sizes from the same study

```{r Multi-level models}
All_richness_multilevel <- rma.mv(yi = yi, V = vi, 
                                  random = ~ 1 | Study_ID/Sample_type, #With this structure we indicate NESTED random effect
                                  #mods = ~Time_since_fire_years-1,
                                  data = All_fungi_richness)
summary(All_richness_multilevel)
```

```{r Orchard and catepillar plots}
## PLOT 
a<- orchaRd:: orchard_plot(sap_alpha_multilevel, mod = "1",
  group="Study_ID", alpha = 0.5, xlab="Hedge's g")+
  #annotate(geom="text", x= 0.80, y= 1, label= paste0("italic(I)^{2} == ", round(I2[1],4), "*\"%\""), 
  scale_fill_manual(values="#dd3497")+
  scale_colour_manual(values="#dd3497")


# We can visualise the results of the meta-analysis using a caterpillar plot
b <- orchaRd::caterpillars(sap_alpha_multilevel, mod="1", 
                      xlab = "Effect size (Hedge's g)",group="Study_ID",
                      colpoint = "grey30",colerrorbar = "#ae017e")

combined <- (b + a)
#ggsave("combined_sap_alpha.png", width = 10, height = 6, dpi = 300)

##################### FIRE SEVERITY #############################

res <- orchaRd::mod_results(sap_richness_multilevel, mod = "Fire_severity", group = "Study_ID")
res
res1 <- orchaRd::mod_results(sap_richness_multilevel, mod="Fire_severity", group = "Study_ID", at=list("Fire_severity" = "Low_severity"), subset = TRUE )
res1
res2 <- orchaRd::mod_results(sap_richness_multilevel, mod="Fire_severity", group = "Study_ID", at=list("Fire_severity" = "Moderate_severity"), subset = TRUE )
res2
res3 <- orchaRd::mod_results(sap_richness_multilevel, mod="Fire_severity", group = "Study_ID", at=list("Fire_severity" = "High_severity"), subset = TRUE )
res3

LS <- orchaRd::orchard_plot(res1, 
                   mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g", alpha = .6) +
  scale_fill_manual(values="#fec44f")+
  scale_colour_manual(values="#fec44f")
MS <- orchaRd::orchard_plot(res2, 
                   mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g", alpha = .6) +
  scale_fill_manual(values="#fd8d3c")+
  scale_colour_manual(values="#fd8d3c")
HS <- orchaRd::orchard_plot(res3,
                  mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g", alpha = .6) +
  scale_fill_manual(values="#ef3b2c")+
  scale_colour_manual(values="#ef3b2c")

combines_sev <- (LS / MS / HS)
#ggsave("sev_sap_richness.png", width = 10, height = 6, dpi = 300)

LS <- orchaRd::caterpillars(res1, 
                   mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g",
                   colpoint = "#fec44f",colerrorbar = "grey30")
MS <- orchaRd::caterpillars(res2, 
                   mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g",
                   colpoint = "#d94801",colerrorbar = "grey30")
HS <- orchaRd::caterpillars(res3,
                  mod = "Fire_severity", group = "Study_ID", xlab = "Hedge's g",
                   colpoint = "#67000d",colerrorbar = "grey30")

combines_sev <- (LS / MS / HS)
ggsave("sev_sap_richness.png", width = 6, height = 5, dpi = 300)

##################### FIRE TYPE  #############################

res <- orchaRd::mod_results(sap_richness_multilevel, mod = "Fire_type", group = "Study_ID", at=list("Fire_type" = "Prescribed_fire"), subset = TRUE )
res
res1 <- orchaRd::mod_results(sap_richness_multilevel, mod="Fire_type", group = "Study_ID", at=list("Fire_type" = "Wildfire"), subset = TRUE )
res1

WF <- orchaRd::orchard_plot(res1, 
                   mod = "Fire_type", group = "Study_ID", xlab = "Hedge's g", alpha = .6) +
  scale_fill_manual(values="#6a51a3")+
  scale_colour_manual(values="#6a51a3")
PF <- orchaRd::orchard_plot(res, 
                   mod = "Fire_type", group = "Study_ID", xlab = "Hedge's g", alpha = .6) +
  scale_fill_manual(values="#1d91c0")+
  scale_colour_manual(values="#1d91c0")

combined_ft <- (WF/PF)
#ggsave("ft_sap_alpha.png", width = 10, height = 6, dpi = 300)


WF <- orchaRd::caterpillars(res1, 
                   mod = "Fire_type", group = "Study_ID", xlab = "Hedge's g",
                   colpoint = "#6a51a3",colerrorbar = "grey30")
PF <- orchaRd::caterpillars(res, 
                   mod = "Fire_type", group = "Study_ID", xlab = "Hedge's g",
                   colpoint = "#1d91c0",colerrorbar = "grey30")

combined_ft <- (WF/PF)
ggsave("ft_sap_richness.png", width = 10, height = 6, dpi = 300)

##################### SOIL DEPTH  #############################


```

```{r}
ggplot(RR_ECM, aes(x = Time_since_fire_months, y = RRbeta)) +
  geom_point(size = 2, color = "darkgreen") +          # points for each study
  geom_smooth(method = "lm", se = TRUE, color = "blue") +  # regression line with CI
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + # reference line
  labs(
    x = "Time since fire (months)",
    y = "Effect size (SMD, Burned − Unburned)",
    title = "Meta-regression: Saprotrophs richness vs time since fire"
  ) +
  theme_minimal(base_size = 14)
```

```{r RR effect sizes categorization}
RR_ECM <- ordinations_df %>%
  dplyr::filter(str_starts(Response_var, "ECM"))

ECM_beta_result <- rma(RRbeta, var_RRbeta, data = RR_ECM) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(ECM_beta_result)

ECM_struc_result <- rma(RRstructure, var_RRstructure, data = RR_ECM) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(ECM_beta_result)

RR_Saprotroph <- ordinations_df %>%
  dplyr::filter(str_starts(Response_var, "Saprotroph"))

RR_Sporocarp <- ordinations_df %>%
  dplyr::filter(str_starts(Response_var, "Sporocarp"))

RR_all_fungi <- ordinations_df %>%
  dplyr::filter(
    !str_starts(Response_var, "ECM") &
    !str_starts(Response_var, "Saprotroph") &
    !str_starts(Response_var, "Sporocarp")
  )

All_beta_result <- rma(RRbeta, var_RRbeta, data = RR_all_fungi) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(All_beta_result)

All_struc_result <- rma(RRstructure, var_RRstructure, data = RR_all_fungi) #yi = EFFECT SIZE; vi = VARIANCE OF THE EFFECT SIZE
summary(All_struc_result)
```

```{r Publication Bias}
funnel(All_struc_result) #we look for symmetry

##Trim and fill analysis
trimandfill <- trimfill(All_struc_result)
#Estimated number of missing studies on the right side: 0 (SE = 10.2953)
trimandfill
funnel(trimandfill)

regtest(All_struc_result)
#Test for Funnel Plot Asymmetry: z = -6.3480, p < .0001 #This SIGNIFICANCE means our funnel plot is ASYMMETRICAL
```

```{r Models: RR}
m_beta <- rma.mv(yi  = RRbeta, V  = var_RRbeta,
                 mods = ~Fire_severity -1, 
                 random = ~ 1 | Study_ID,
                 data = RR_all_fungi)
summary(m_beta)

m_structure <- rma.mv(yi  = RRstructure, V  = var_RRstructure,
                 #mods = ~Fire_type-1, 
                 random = ~ 1 | Study_ID,
                 data = RR_all_fungi)
summary(m_structure)

```

```{r Forest plot}
png("All_beta_result_plot.png", width = 2000, height = 3000, res = 300)
forest.rma(ECM_struc_result, slab = RR_ECM$Study_ID, header="study")
dev.off()
```







